---
title: "Exploring Palmer Penguins with R"
author: "Yuvraj Singh Walia"
date: "2026-09-23"
format: html
---

## Introduction

In this post, I explore the Palmer Penguins dataset using R. I will compare
body mass across penguin species and investigate how bill measurements differ
between species.

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
Palmer Station Antarctica LTER.

The Palmer Penguins dataset was collected by Dr. Kristen Gorman and the
Palmer Station Long Term Ecological Research program. The dataset is available
under the CC0 licence.

## Loading the data

I load the Palmer Penguins package and inspect the first few observations.

```{r}
library(palmerpenguins)

penguins <- palmerpenguins::penguins
head(penguins)
```

## Understanding the dataset

I first check the size of the dataset and count the missing values in each
variable.

```{r}
cat("Number of observations:", nrow(penguins), "\n")
cat("Number of variables:", ncol(penguins), "\n")

colSums(is.na(penguins))
```

The dataset contains measurements for individual penguins from three species.
Some variables contain missing observations, so these need to be considered
before calculating summaries.

## Preparing the data

For this analysis, I focus on species, body mass, bill length, and bill depth.
I remove rows that have missing values in any of these variables.

```{r}
analysis_penguins <- penguins[
  complete.cases(
    penguins[, c("species", "body_mass_g", "bill_length_mm", "bill_depth_mm")]
  ),
  c("species", "body_mass_g", "bill_length_mm", "bill_depth_mm")
]

cat("Rows before removing missing values:", nrow(penguins), "\n")
cat("Rows after removing missing values:", nrow(analysis_penguins), "\n")

head(analysis_penguins)
```

Removing incomplete observations ensures that the summaries and plots below
are based on penguins with all of the measurements required for this analysis.

## Comparing species

Next, I calculate the average body mass, bill length, and bill depth for each
penguin species.

```{r}
species_summary <- aggregate(
  cbind(body_mass_g, bill_length_mm, bill_depth_mm) ~ species,
  data = analysis_penguins,
  FUN = mean
)

species_summary[, -1] <- round(species_summary[, -1], 1)

species_summary
```

The summary shows that the three species differ in their typical body
measurements. Gentoo penguins have the greatest average body mass, while the
species also show noticeable differences in average bill length and bill depth.

## Body mass by species

To see the differences in body mass more clearly, I create a boxplot for the
three penguin species.

```{r}
#| label: fig-body-mass-species
#| fig-cap: "Distribution of body mass across the three Palmer penguin species."

boxplot(
  body_mass_g ~ species,
  data = analysis_penguins,
  xlab = "Species",
  ylab = "Body mass (g)",
  main = "Body mass by penguin species"
)
```

The figure shows clear differences in body mass among the species. Gentoo
penguins tend to have higher body mass values than Adelie and Chinstrap
penguins.

## Relationship between bill measurements

Finally, I calculate the correlation between bill length and bill depth
separately for each species.

```{r}
bill_correlations <- sapply(
  split(analysis_penguins, analysis_penguins$species),
  function(group) {
    cor(group$bill_length_mm, group$bill_depth_mm)
  }
)

round(bill_correlations, 3)
```

These values show how bill length and bill depth are related within each
species. Calculating the correlations separately avoids mixing the three
species together when examining this relationship.

## Conclusion

This analysis used R to explore differences in body and bill measurements
among Palmer penguin species. After checking for missing data and preparing a
complete dataset, I compared average measurements across species and
visualized the distribution of body mass.

The results show clear differences among Adelie, Chinstrap, and Gentoo
penguins. Gentoo penguins are heavier on average, and the species also differ
in their bill measurements. The analysis demonstrates how data cleaning,
summary statistics, visualization, and correlation can be used to explore a
biological dataset in R.